---
title: "Repeated measures intensity"
output: html_document
editor_options: 
  chunk_output_type: console
bibliography: References4.bib
---

# Low- vs high-repetition training for strength development

## Introduction
What is the recommended intensity for strength development, low- or high-repetition training? Anderson & Kearney [-@Anderson1982] conducted a study on male college students, and found that after a 9 weeks training intervention, 3x6-8 repetition maximum (RM) training, gave a greater increase in 1 RM benchpress compared to 2x30-40 RM training (20 % and 8 %, respectively). However, Stone & Coulter [-@Stone1994] who conducted a similar 9 weeks training intervention on female collage students, found that both 3x6-8 RM and 1x30-40 RM gave significant pre to post increases in 1 RM in both upper and lower body exercises. In the latter study they could not find a significant difference between the groups after the training intervention. These results suggests that both low- and high-repetition training improves strength measured as 1 RM. To find out if one of the methods are superior to the other and therefore should be recommended, more studies are required.

As a result of this, we wanted to take a closer look at the effects of low- and high-repetition training. The goal of the study is to find out which training intensity, 10 or 30 RM, gives the larger development of strength measured as 1 RM legpress.

#### Question
> What should be the recommended intensity for strength development?

#### Hypothesis
> 10 RM gives a larger development of strength measured as 1 RM legpress, compared to 30 RM

## Methods
#### Study design
To compare the effects of 10 and 30 RM training, the participants of the study were assigned to one of two groups: the 10 RM group or the 30 RM group. Strength was tested as 1 RM legpress and benchpress before (pre), during (mid) and after (post) the training intervention. 27 participants completed the first two tests, and 24 participants completed all three tests.

#### Statistical analysis
We wanted to estimate the difference between the 10 RM and the 30 RM group in 1 RM legpress when controlling for baseline after a strength training intervention. The data set contained several variables, so we first had to reduce our data to only contain our variables of interest: the 1 RM load in legpress at pre-, mid- and post-test for both the 10 RM and 30 RM group. Because we have repeated measures, we have to account for correlated data to avoid that it affects the statistical power. A mixed model accounts for potential differences between the groups at baseline, and was therefore used to estimate the difference in mean between the two groups at both pre-, mid- and post-test. Confidence intervals were used to evaluate the effects. The data was plotted and presented in a combined plot, where the first one shows changes in strength for each participant and estimated averages in each group at time-points pre, mid and post, and the second shows the average group differences at time-points mid and post (figure 1). The statistics of the estimates is presented in a table (table 1).


## Results
```{r, include=FALSE, warning=FALSE, message=FALSE}
# Chunck to run the code

# Loading needed packages
library(tidyverse)
library(readxl)
library(kableExtra)
library(knitr)
library(broom)
library(lme4)
library(emmeans)
library(cowplot)

# Loading and checking the data
data <- read_xlsx("./data/ten_vs_thirty.xlsx", na = "NA") 

# Preparing the data for legpress
leg.data <- data %>% 
  filter(!is.na(load)) %>% # Rempving NA from the variable "load"
  filter(exercise == "legpress",           # Chosing legpress
         timepoint %in% c("pre", "mid", "post")) %>%  # Choosing timepoints
  mutate(timepoint = factor(timepoint, 
                            levels = c("pre", "mid", "post")),# Fixing the order of timepoint)
         group = factor(group,
                        levels = c("RM30", "RM10"))) # Fixing the order of groups (comparing to 30 RM)

leg.data # Printing the results

# Plotting a figure that shows changes in strength for each participant in each group
leg.figure <- leg.data %>%
  mutate(timepoint = factor(timepoint, levels = c("pre", "mid", "post"))) %>%
  ggplot(aes(timepoint, load, group = subject, color = group)) + # Deciding axis and adding a color to each group
  geom_line() + # Each subject gets a line
  labs(x = "Time-point", y = "1 RM (kg)") +    # Changing the axis´ names
  theme_minimal()                              # Chosing layout

leg.figure # Printing the figure
# Figure: Changes in strength for each participant in both the 10 RM and 30 RM group

# Fitting a mixed model
# Want to compare the two groups and model any differences in change over time.
# A mixed model can account for repeated data by using the subject indicator as a random effect.
# Letting the groups vary from each other with timepoint * group. Let them vary over the timepoints
# Each participant will be compared from their own starting point (intercept).
lmer1 <- lmer(load ~ timepoint * group + (1|subject), data = leg.data)


summary(lmer1) # Printing the results

## Fixed effects from the lmer1 model
# Intercept are the average at pre (10 RM)
# Timepointmid are the difference pre to mid (10 RM)
# Timepointpost are the difference pre to post (10 RM)

# GroupRM10 are the difference between 10 RM and 30 RM at pre
# Timepointmid:groupRM10 are the difference between 10 RM and 30 RM at mid
# Timepointpost:groupRM10 are the difference between 10 RM and 30 RM at post

## Variables of interest
# The difference between the groups when we have controlled for baseline after a trainingintervention
# 11 kg difference between 10 RM and 30 RM at mid 
# 38kg difference between 10 RM and 30 RM at post



# Residual plot to check if our assumptions are met
plot(lmer1)
# The residual variation is similar over the range of fitted values, so this is an ok model we can use.


# Model estimates
# Estimated averages at each time-point within each group
est <- emmeans(lmer1, specs = ~ timepoint|group) # Specs specifies what means to calculate

est # Printing the results


# Plotting the model estimates (estimated averages)
# Have to convert the object created with emmeans to a data frame to be able to plot it
est.figure <- est %>%
  data.frame() %>% 
  mutate(timepoint = factor(timepoint, levels = c("pre", "mid", "post"))) %>%
  ggplot(aes(timepoint, emmean, group = group, color = group)) + 
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), # Represent the 95% CI
                position = position_dodge(width = 0.2), # Moves the errorbars away from each other, same with the lines and points further down
                width = 0.1) + # Decides the width of the errorbars, the same under
  geom_line(position = position_dodge(width = 0.2)) +
  geom_point(position = position_dodge(width = 0.2)) +
  labs(x = "Time-point", y = "1 RM (kg)") + # New labels on the axis
  theme_minimal()
# Figure: Estimated averages in each group at time-points pre, mid and post

est.figure # Printing the figure

## Tips!
# Choosing between SD og CI, CI are preferred when plotting
# Remember to only present a result once! Not the same result in both a table and a figure!


raw.est.figure <- est %>%
  # The model estimates 
  data.frame() %>%
  mutate(timepoint = factor(timepoint, 
                            levels = c("pre", "mid", "post"))) %>%
  ggplot(aes(timepoint, emmean, group = group, color = group) ) + 
  
  # Adding raw data
  geom_line(data = leg.data, aes(timepoint, load, group = subject, color = group), 
            alpha = 0.4) + # Add transparency to individual lines
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
                position = position_dodge(width = 0.2), 
                width = 0.1) +
  geom_line(position = position_dodge(width = 0.2)) +
  geom_point(position = position_dodge(width = 0.2)) +
  labs(x = "", y = "1RM (kg)", 
       color = "Intensity") + # New labels on the axis
  # Removing text and indicators on the x-axis as we are going to combine this plot with another, and then we use the text and indicators in that plot
  # FUNGERER IKKE
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank()) +
  theme_minimal()
# Figure: Changes in strength for each participant and estimated averages in each group at time-points pre, mid and post


raw.est.figure # Printing the figure



# Saving the confidence intervals
# Getting the 95% CI from each model coefficient
conf_intervals <- confint(lmer1)
conf_intervals
# CI at post do not contain 0. Ergo, we have proof against the null hypothesis that says that there are no difference between the groups. There is a difference!
# Present random effects as well?

# Saving the regression coefficients
# Getting the estimates of the interaction coefficients from the model
coefs <- summary(lmer1)$coef
coefs

# Using cbind (column bind) to combine the two data frames into one data frame
coef_summary <- cbind(coefs, data.frame(conf_intervals)[3:8, ]) 
coef_summary

# Printing a table using kable()
coef.table <- coef_summary %>%
  kable(col.names = c("Estimate", "Std. Error", "t-value", "Lower CI", "Higher CI"), # Column names
        digits = c(1, 1, 2, 1, 1),# Deciding how many decimals at each value
        caption = "**Table 1:** Coefficient estimates, standard error, t-statistic, lower and upper confidence interval at time-points pre, mid and post in both the 10 RM and 30 RM group.")
# Table: Coefficient estimates, standard error, t-statistic, lower and upper confidence interval

coef.table   # Printing the table

# HVORDAN FÅ DETTE TIL Å BLI BEDRE? 
# Ta med hele tabellen eller bare de to siste radene?
# Hvordan få tabellen og tabellteksten lenger og svart?
# Hvordan få mellomrom mellom resultatteksten og tabellen?



# Plot to show "What should be the recommended intensity for strength development?"
# Interested in the estimates "timepoint:groupRM10" and "timepointpost:groupRM10"

# Reducing the data frame and plotting it to show estimated differences between groups at mid and post after accounting for baseline differences
est.diff.figure <- coef_summary %>%
  mutate(coef = rownames(.)) %>%
  # Filter only the interaction variables
  filter(coef %in% c("timepointmid:groupRM10", "timepointpost:groupRM10")) %>%
  # Making a "timepoint" variable to represent the "leg.data" data set.
  mutate(timepoint = gsub("timepoint", "", coef), 
         timepoint = gsub(":groupRM10", "", timepoint)) %>%
  # Adding a row so that timepoint = "pre" is represented
  add_row(timepoint = "pre", coef = "pre") %>%
  # Fixing the order of the time variable
  mutate(timepoint = factor(timepoint, levels = c("pre", "mid", "post"))) %>%
  # Creating the plot
  ggplot(aes(timepoint, Estimate)) + 
  # Adding a line indicating zero geom_hline (horizontal line)
  geom_hline(yintercept = 0, lty = 2) +
  # Adding errorbars and points
  geom_errorbar(aes(ymin = X2.5.., ymax = X97.5..), width = 0.1) +
  geom_point(shape = 24, size = 3, fill = "white") +
  # Changing axis titles and title in the legend
  labs(x = "Time-point", y = "Average group differences\n1RM (kg, 95% CI)", 
       color = "Intensity") + 
  theme_minimal()

est.diff.figure # Printing the figure
# Figure: Average group differences in 1 RM legpress at mid and post, comparing 10 RM to 30 RM.

# Answer to our question: 10 RM should be the recommended intensity
# The points show the difference between the groups at the different time-points, comparing 10 RM to 30 RM.


# Combining the plots to one plot
# Using "raw.est.figure" and "est.diff.figure"
figure1 <- plot_grid(raw.est.figure, est.diff.figure, ncol = 1, 
          # Aligning vertically so that the x axis matches
          align = "v", 
          # Setting alignment so that axis align both on the left (l) and right (r) axis
          axis = "lr")

figure1 # Printing the figure
# Figure 1: Changes in 1 RM legpress for each participant and estimated averages in each group at time-points pre, mid and post. Average group differences in 1 RM legpress at time-points mid and post.


# Summary statistics, getting even more information about the model we have fitted
summary(lmer1)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.keep='first', fig.cap= "**Figure 1:** Changes in 1 RM legpress for each participant and estimated averages in the 10 RM and 30 RM group at time-points pre, mid and post. Average group differences in 1 RM legpress at time-points mid and post, comparing 10 RM to 30 RM."}
# Chunk to print the results


figure1 # Printing the figure


```

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.keep='first', fig.cap= "**Table 1:** Coefficient estimates, standard error, t-statistic and lower and upper confidence intervals at time-points pre, mid and post in both the 10 RM and 30 RM group."}
# Chunk to print the results

coef.table # Printing the table

```

The 10 RM group lifted 7.6 (27.9) kg less than the 30 RM group at pre (baseline). At mid and post, the 10 RM group lifted 11.1 (12.5) kg and 38.6 (13.0) kg more than the 30 RM group, respectively.

## Discussion
As the confidence intervals contained 0, there were no difference between the groups in 1 RM legpress at pre (95% CI: [-62.0, 46.8], *t*(26) = -0.27), nor at mid (95% CI: [-12.8, 35.0], *t*(26) = 0.89), respectively. At post-test after the training intervention, there was a difference between the 10 RM group and the 30 RM group, as the confidence interval did not contain 0 (95% CI: [13.6, 63.6], *t*(23) = 2.96). 

This study therefore suggests that 10 RM training gives a larger development of strength measured as 1 RM in legpress, compared to 30 RM training. Our findings indicate that 10 RM should be the recommended intensity for strength development. That high-intensity training may be superior to low-intensity training, is supported by the findings of Anderson & Kearney [-@Anderson1982]. 

## References